### 2023-06-22 submission/manuscript.tex
Submitted

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font

@@

@@

\hypertarget{rarefy-your-data}{%
\section{Rarefy your data}\label{rarefy-your-data}}

@@

\textbf{Running title:} Rarefy your data

@@

\({\dagger}\) To whom corresponsdence should be addressed:

@@

@@

The ability to generate millions sequence reads from 16S rRNA gene
amplicons and metagenomic datasets has allowed researchers to multiplex
multiple samples on the same sequencing run by pooling separate PCRs
that can be deconvoluted later based on index (aka barcode) sequences
that embeded into the primer sequence (1, 2). Unfortunately, it is
common to observe variation in the number of sequence reads from each
sample vary by as much as 100-fold (e.g, see Figure S1). This occurs
because pooling of DNA from multiple PCRs is frought with numerous
opportunities for technichal errors to compound leading to a skewed
distribution. Aside from developing better methods of pooling DNA, the
question of how to control for uneven sampling effort in microbial
ecology studies has become controversial.

The practice of rarefaction has been commonly used in ecology for more
than 50 years as a tool to control for uneven sampling effort across
experimental replicates (3, 4). Microbial ecologists have used it to
compare 16S rRNA gene sequence data for the past 25 years (5--7). When
data are rarefied, the investigator selects a desired threshold of
sampling effort and removes any samples below that threshold. They then
randomly select that many sequences, with replacement from each sample.
Based on the obseved sequence counts, the researcher can then calcuate
alpha diversity metrics including richness and diversity indices or beta
divesity metrics such as a Jaccard or Bray-Curtis dissimilarity index. I
refer to this single sampling as a subsample; this method is implemented
as the sub.sample function in mothur (8) and the rrarefy function in the
vegan R package (9). Rarefaction repeats the subsampling a large number
of times (e.g., 100 or 1,000) and calculates the mean of the alpha or
beta diersity metric over those subsamplings; rarefaction is implemented
in mothur using the summary.single and dist.shared functions (8) and
with the vegan R package using the rarefy or avgdist fuctions (9).
Rarefaction effectively tells a researcher what an alpha or beta
diversity metric would have been for a collection of samples if they
were all evenly sampled. Although a closed form equation exists to
calcualte the expected richness (4), it is computationally easier to
empirically calculate rarefied richness and other alpha and beta
diversity metrics.

In 2014, McMurdie and Holmes (10) announced that rarefaction of
microbial community data is ``statistically inadmissible'' because it
omits valid data. In their simulations, they observed that the use of
rarefaction reduced the statistical power to compare communities based
on beta diversity metrics. The communities used in this study were
simulated by mixing operational taxonomic unit (OTU) frequencies from
marine and human fecal communities, which were then sampled to generate
40 replicate communities each sampled to different depths. The
simulation with the greatest depth of coverage obtained an average of
10,000 sequences per sample. Analysis of the R code that they used
indicates that the simulations were actually only based on a single
subsample and not truly rarefied data. Furthermore, the authors removed
any OTU from the simulated communities that was not observed in three or
more samples. This is a frequently used approach for analyzing
microbiome data (11). Unfortunately, it removes valid data, skews the
distribution of the community, and removes true patchiness from the
representation of the communities (12). Finally, the analysis did not
include analysis of alpha diersity metrics and chose to assess beta
diversity metrics based on clustering quality of samples rather than
using the more frequently used approach of employing non-parametric
multivariate analysis of variance (13). Regardless of these caveats,
McMurdie and Holmes recommended the use of variance stabilizing
transformations such as those applied for gene expression analysis such
as that found in the DeSeq R package.

Beyond the critiques from McMurdie and Holmes others have developed
alternative approaches to controlling for uneven sampling effort in
amplicon sequencing studies. For alpha diversity metrics, Willis used
toy datasets to demonstrate that one could estimate the richness for
each sample in a dataset and use those values for statistical
comparisons (14). Non-parametric estimators of richness and diversity
(15, 16) and parametric estimators of richness {[}Willis2015{]} that
have been used in microbial ecology studies. For beta diversity metrics
at least four approaches have been pursued. First, one could use
relative abundance values where the observed number of sequences in an
OTU is divided by the total number of sequences in the sample (17).
Second, normalization strategies have been developed where the relative
abundance is mutiplied by the size of the minimum desired sampling
effort and fractional values are reapportioned to the OTUs to obtain
integer values (18, 19). Third, a variety of center log-ratio methods
have been developed where the compositional nature of the OTU counts is
removed and used to calculated Euclidean distances (aka Aitchison
distances) {[}(20); (19); (17); (21); (22). This strategy is purported
to control for uneven sampling effort (21, 23); however, some have
noticed that this feature breaks down under certain conditions (24).
Finally, as mentioned above, variance stabilization transformations have
been recommended to generate values that can be used to calcualte
Euclidean distances (10).

The ongoing controversy over the use of rarefaction and the development
of numerous strategies to control for uneven sampling effort caused me
to question how these methods compared to each other. My analysis
included 16S rRNA gene sequence data from from 12 studies that
characterized the variation in bacterial communities from diverse
environments (Table 1 and Figure S1). The sequences were assigned to
OTUs using a standard pipeline and their frequencies and the number of
sequences found in each sample were used to generate simulated
communities and treatment effects. For each dataset and simulation, 100
replicate datasets were generated and used as input to each of the
strategies for controlling for uneven sampling effort. My overall
conclusion was that rarefaction outperformed the alternative strategies.

@@

\textbf{\emph{Without rarefaction, metrics of alpha diversity are
sensitive to sampling effort.}} To test the sesitivity of various
approaches of measuring alpha diversity to sampling effort, I generated
null models for each dataset. Under a null model, each community from
the same dataset would be expected to have the same alpha diversity
regardless of the sampling effort. I measured the richness of the
communities in each dataset without any correction, using scaled ranked
subsampling (SRS) normalized OTU counts, with estimates based on
non-parametric and parametric approaches, and using rarefaction
(e.g.~Figure S2). For each dataset, all of the approaches, except for
rarefaction, showed a strong correlation between richness and the number
of sequences in the sample (Figure 1A). Next, I assessed diversity using
the Shannon diversity index and the inverse Simpson diversity index
without any correction, using normalized OTU counts, and rarefaction; I
also used a non-parametric estimator of Shannon diversity. The
correlation between sampling depth and the diversity metric was not as
strong as it was for richness and the inverse Simpson diversity values
were less sensitive than the Shannon diversity values; however, the
correlation to the rarefied diversity metrics were the lowest for all of
the metrics and studies (Figure 1A). The rarefied alpha-diversity
metrics consistently demonstrated a lack of sensitivity to sampling
depth.

\textbf{\emph{Without rarefaction, metrics of beta diversity are
sensitive to sampling effort.}} To test the sesitivity of various
approaches of measuring beta diversity to sampling effort, I used the
same null models used for studying the sensitivity of alpha diversity.
Under a null model, the ecological distance between any pair of samples
would be the same regardless of the difference in the number of
sequences observed in each sample (e.g., Figure S3). First, I calculated
the Jaccard distance coefficient between all pairs of communities within
a dataset. The Jaccard distance coefficient is the fraction of OTUs that
are unique to either community and does not account for the abundance of
the OTUs. Jaccard distances were calcualted using the uncorrected OTU
counts, with rarefaction, relative abundances, and following
normalization using cumulative sum scaling (CSS) and SRS. Only the
rarefied distances showed a lack of sensitivity to sampling effort
(Figure 1B). Second, I analyzed the sensitivity of the Bray-Curtis
distance coefficient, which is a popular metric that incorporates the
abundance of each OTU. Similar to what I observed with the Jaccard
coefficient, only the rarefied data showed a lack of sensitivity to
sampling effort (Figure 1B). Third, I calcualted the Euclidean distance
on raw OTU counts where the central log-ratio (CLR) was calculated
(i.e., Aitchison distances) by ignoring OTUs in samples with zero counts
(Robust CLR), adding a pseudocount of one to all OTU counts prior to
calculating the CLR (One CLR), adding a pseudocount of one divided by
the total number of sequences obtained for the community (Nudge CLR),
and imputing the value of zero counts (Zero CLR). The Aitchison
distances were all strongly sensitive to sampling effort (Figure 1B).
Finally, I used the variance-stabilization transformation (VST) from
DeSeq2 prior to calculating Euclidean distances. Again, there was a
strong sensitivity to sampling effort (Figure 1B). Although Euclidean
distances are not typically used on raw or rarefied count data in
ecology, rarefied Euclidean distances were not sensitive to sampling
effort. Across each of the beta diversity metrics and approaches used to
account for uneven sampling effort and sparsity, rarefaction was the
least sensitive approach to differences in sampling effort.

\textbf{\emph{Rarefaction limits the detection of false positives when
sampling effort and treatment group are confounded.}} Next, I
investigated the impact of the various strategies and metrics on falsely
detecting a significant difference using the the same communities
generated from the null model in the analysis of alpha and beta
diversity metrics. To test for differences in alpha and beta diversity I
used the non-parametric Wilcoxon test and non-parametric
permutation-based multivariate analysis of variance (PERMANOVA). First,
I employed an unbiased null treatment model to measure the false
detection rate, which should not have meaningfully differed from 5\%.
Indeed, for each dataset and alpha and beta diversity metric and
strategy for accounting for uneven sampling, approximately 5\% of the
tests yielded a significant result (Figure 2). Second, I employed a
biased null treatment model where the treatment group was completely
confounded with the number of sequences in each sample. Under this
model, only the rarefied data consistently resulted in a 5\% false
positive rate for alpha and beta diversity metrics (Figure 2). These
results aligned with the observed sensitivity of alpha and beta
diversity metrics to sampling effort and underscore the value of
rarefaction.

\textbf{\emph{Rarefaction preserves the statistical power to detect
differences between treatment groups.}} To assess the impact of
different approaches to control for uneven sampling effort I performed
two additional simulations. In the first simulation, I implemented a
skewed abundance distribution model to create two treatment groups for
each datasets that were each populated with half of the samples each
with the same number of sequences as the samples had in the observed
data. The power to detect differences in richness between the two
simulated treatment groups by all approaches was low (Figure 4A). This
was likely because the approach for generating the perturbed community
did not necessarily change the number of OTUs in each treatment group.
Regardless, the simulations testing differencse in richness using the
Rice and Stream datasets had the greatest power when the richness data
were rarefied. To explore this further, a richness-adjusted community
model was created by removing 3\% of the OTUs from a null model model.
As suggested by the first simulation, the rarefied richness data had a
higher statistical power than the other approaches when measuring
richness (Figure 5). The simulations testing the power to detect
differences in Shannon diversity also showed that rarefied data
performed other methods (Figure 4A). When testing for differences in the
Inverse Simpson diversity index the the difference between rarefaction
and the other methods was negligible (Figure 4A). For tests of beta
diversity I found that rarefaction was the most reliable approach to
maintain statistical power to detect differences between two
communities. Among the tests using the Jaccard and Bray-Curtis metrics,
raw count data and CSS normalized data had little power relative to
rarefied, relative abundance, and SRS normalized data. The differences
in power between rarefied, relative abundance, and SRS normalized data
was small, but if there were differences, the power obtained using
rarefied data was greater than the other methods. Among the tests using
Euclidean distances, using raw counts and CLR and DeSeq2 transformed
data had little power relative to the distances calcualted using
rarefied and relative abundance data. This power-based analysis of the
simulated communities using different methods of handling uneven sample
sizes demonstrated the value of rarefaction for preserving the
statistical power to detect differences between treatment groups for
measures of alpha and beta diversity.

\textbf{\emph{Increased rarefaction depth reduces intra-sample variation
in alpha and beta diversity.}} Once concern with rarefying communities
is the perceived loss of sequencing information when more a large
fraction of data appears to be removed when the community with the
greatest sequencing depth is rarefied to the size of the community with
the least (e.g., 99\% with the Bioethanol dataset). To assess the
sensitivity of alpha and beta diversity metrics to rarefaction depth, I
again used the dataset generated using the null models, but rarefied
each community to varying sampling depths (Figure 6). The richness
values increased with sampling effort as rare OTUs would continue be
detected. In contrast, the Shannon diversity and Bray-Curtis values
plateaued with increased sampling effort. This result was expected since
increased sampling would lead to increased precision in the measured
abundance of OTUs. Next, I measured the coefficient of variation (i.e.,
the mean divided by the standard deviation) between samples for
richness, Shannon diversity, and Bray-Curtis distances. Although the
richness values appeared to increased unbounded with smapling effort,
the coefficient of variation for each dataset was relatively stable. In
general, the coefficient of variation increased slightly with sampling
depth only to decline once smaller samples were removed from the
analysis at higher sampling depths. Interestingly, the coefficient of
variation between Shannon diveristy values decreased towards zero with
increased sampling effort and the coefficient of variation between
Bray-Curtis distances tended to increased. Regardless, the coefficients
of variation were relatively small.

\textbf{\emph{Most studies have a high level of sequencing coverage.}}
To explore the concern over loss of sequencing depth further, I
calculated the Good's coverage for the observed data. The median
coverage for each dataset ranged between 89.4 and 99.8\% for the
Seagrass and Human datasets, respectively (Figure 7). When I rarefied
each dataset to the size of the smallest community in the dataset, with
the exception of the Seagrass, Rice, and Stream datasets, the median
coverage for the rarefied communities was still greater than 90\%. These
results suggest that most studies had a level of sequencing coverage
that aligned with the diversity of the communities. Next, I used the
null model for each dataset to ask how much sequencing effort was
required to obtain higher levels of coverage. To obtain 95 and 99\%
coverage, an average of 2.70 and 101.20-fold more sequence data was
estimated to be required than was required to obtain 90\% coverage,
respectively (Figure 7). As suggested by the simulated coverages curve
in Figure 7, these estimates are conservative. Regardless, the
sequencing effort required to acheive higher sequencing depth would
likely limit the number of samples that could be sequenced when
controlling for costs. Although it may be disconcerting to rarefy to a
sequencing depth that is considerably lower than that obtained for the
best sequenced community in a dataset, sequencing coverage for many
environments is probably adequate even at the lower sequencing depth. Of
course, the results above have demonstrated that rarefaction is
necessary to avoid problems with making inferences.

@@

Over the past decade, the question of whether to rarefy microbial
community sequence data has become controversial. The analyses I
presented here strongly indicate that rarefaction is necessary to
control for uneven sampling effort when comparing communities based on
alpha and beta diversity indices. Compared to all other methods,
rarefaction removed the correlation between sequencing depth and alpha
or beta diversity metrics when the sequencing depth varied by as much as
97-fold across samples. I showed that this correlation could lead
falsely detecting differences between treatment groups if sampling depth
and sequencing effort are confounded (e.g., 25). The correlation with
sampling effort leads to an artificial increase in the variation between
samples and a reduced power to detect true differences in alpha and beta
diversity. For these reasons, rarefaction is a valuable tool to control
for uneven sampling effort until improved statistical procedures are
developed or it becomes possible to more evenly distribute sequencing
effort across samples.

The primary alternative to rarefaction that many advise is to estimate
alpha diversity has been to use non-parametric or parametric methods
using raw data and to then compare the estimates (14). My data
demonstrates that such approaches are limited for several reasons.
First, it has been widely observed that non-parametric richness esimates
such as ACE and Chao1 are sensitive to sampling effort. Therefore, these
strategies do not, in practice, remove the effects of sampling effort.
Second, parametric approaches, such as those implemented in the
breakaway R package, generate confidence intervals that are likely to
include the true richness and theoretically shrink with increased
sampling effort. For most samples, the confidence intervals around the
estimates are too wide to be meaningful, again leading to an inability
to remvoe the effects of sampling effort. Third, it has become an
increasingly common practice for researchers to remove sequences that
are rare in a sample (e.g., those that appear once). Although that
approach was not taken in this study, removing rare sequences would skew
the distribtion of sequences and OTUs leading to a distortion of the
measurement of alpha diversity (12). The effects of removing rare
sequences would vary across samples depending on the number of sequences
in each sample. One interesting result of this analysis was the
demonstration that as metrics that depend less on rare taxa are used,
the effect of uneven sampling effort was reduced. For example, richness
counts a sequence appearing once as much as sequence appearing 1,000
times, while Shannon diversityindex places more emphasis on the more
abundant sequence, and the inverse Simpson index even more. Although
normalizing communities to a common number of sequences is also
suggested (e.g., SRS normalization) to control for uneven sampling
effort, the current analysis indicates that its performance does not
meet that of rarefaction. For analysis of alpha diversity metrics,
rarefaction is the most effective and consistent approach to control fo
uneven sampling effort.

>>>>

\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}

@@

\usepackage[document]{ragged2e}

@@

\raggedright

\hypertarget{rarefaction-is-currently-the-best-approach-to-control-for-uneven-sequencing-effort-in-amplicon-sequence-analyses}{%
\section{Rarefaction is currently the best approach to control for
uneven sequencing effort in amplicon sequence
analyses}\label{rarefaction-is-currently-the-best-approach-to-control-for-uneven-sequencing-effort-in-amplicon-sequence-analyses}}

@@

\textbf{Running title:} Use rarefaction

@@

\({\dagger}\) To whom correspondence should be addressed:

@@

Considering it is common to find as much as 100-fold variation in the
number of 16S rRNA gene sequences found across samples in a study,
researchers need to control for the effect of uneven sequencing effort.
How to do this has become a contentious question. Some have argued that
rarefying or rarefaction is ``inadmissible'' because it omits valid
data. A number of alternative approaches have been developed to
normalize and rescale the data that purport to be invariant to the
number of observations. I generated community distributions based on 12
published datasets where I was able to assess the ability of multiple
methods to control for uneven sequencing effort. Rarefaction was the
only method that could control for variation in uneven sequencing effort
when measuring commonly used alpha and beta diversity metrics. Next, I
compared the false detection rate and power to detect true differences
between simulated communities with a known effect size using various
alpha and beta diversity metrics. Although all methods of controlling
for uneven sequencing effort had an acceptable false detection rate when
samples were randomly assigned to two treatment groups, rarefaction was
consistently able to control for differences in sequencing effort when
sequencing depth was confounded with treatment group. Finally, the
statistical power to detect differences in alpha and beta diversity
metrics was consistently the highest when using rarefaction. These
simulations underscore the importance of using rarefaction to normalize
the number of sequences across samples in amplicon sequencing analyses.

@@

The ability to generate millions of 16S rRNA gene amplicon and
metagenomic sequence reads has allowed researchers to multiplex multiple
samples on the same sequencing run by pooling separate PCRs that can be
deconvoluted later based on index (aka barcode) sequences that are
embedded into the sequence of the PCR primers (1--3). Unfortunately, it
is common to observe variation in the number of sequence reads from each
sample vary by as much as 100-fold (e.g, see Figure S1). This occurs
because pooling of DNA from multiple PCRs is fraught with numerous
opportunities for technical errors to compound leading to a skewed
distribution. Aside from developing better methods of pooling DNA, the
question of how to control for uneven sequencing effort in microbial
ecology studies has become controversial.

The practice of rarefaction has been commonly used in ecology for more
than 50 years as a tool to control for uneven sequencing effort across
experimental replicates (4, 5). Microbial ecologists have used it to
compare 16S rRNA gene sequence data for the past 25 years (6--8). With
rarefaction the investigator selects a desired threshold of sequencing
effort and removes any samples below that threshold. They then randomly
select that many sequences, with replacement from each sample. Based on
the observed sequence counts, the researcher can then calculate alpha
diversity metrics including richness and diversity indices or beta
diversity metrics such as a Jaccard or Bray-Curtis dissimilarity index.
I refer to this single sampling as a subsample; this method is
implemented as the \texttt{sub.sample} function in mothur (9) and the
\texttt{rrarefy} function in the \texttt{vegan} R package (10).
Rarefaction repeats the subsampling a large number of times (e.g., 100
or 1,000 times) and calculates the mean of the alpha or beta diversity
metric over those subsamplings; rarefaction is implemented in mothur
using the \texttt{summary.single} and \texttt{dist.shared} functions (9)
and with the \texttt{vegan} R package using the \texttt{rarefy} or
\texttt{avgdist} functions (10). Rarefaction effectively tells a
researcher what an alpha or beta diversity metric would have been for a
collection of samples if they were all sequenced to the same depth.
Although a closed form equation exists to calculate the expected
richness (5), it is computationally easier to empirically calculate
richness and other alpha and beta diversity metrics by rarefaction.

In 2014, McMurdie and Holmes (11) announced that ``rarefying'' of
microbial community data was ``statistically inadmissible'' because it
omits valid data. In their simulations, they observed that rarefying
reduced the statistical power to correctly cluster samples into the same
treatment groups based on beta diversity metrics. Although the detail
was lost on many subsequent researchers, they did not describe
rarefaction, but rarefying (12). According to McMurdie and Holmes,
rarefying is a single subsampling of an OTU abundance table whereas
rarefaction repeats the subsampling step many times. Furthermore, their
simulations penalized the rarefied data by removing 15\% of the samples
and evaluated clustering quality using a clustering algorithm that
performed worse for rarefied data. According to my reanalysis of
rarefaction using the original simulation framework, rarefaction
outperformed other normalization method being used in 2014 for both
alpha and beta diversity metrics (12). Others have also critiqued the
original work (13). Regardless, there is a general sense in the
community that rarefying and rarefaction should not be used with
microbiome data.

Since McMurdie and Holmes published their simulations, others have
developed alternative approaches to control for uneven sequencing effort
in amplicon sequencing studies. For alpha diversity metrics, Willis used
toy datasets to demonstrate that one could estimate the richness for
each sample in a dataset and use those values for statistical
comparisons (14). Non-parametric estimators of richness and diversity
(15, 16) and parametric estimators of richness (17) have been used in
microbial ecology studies. For beta diversity metrics at least four
approaches have been pursued. First, one could use relative abundance
values where the observed number of sequences in an OTU is divided by
the total number of sequences in the sample (18). Second, normalization
strategies have been developed where the relative abundance is
multiplied by the size of the minimum desired sequencing effort and
fractional values are reapportioned to the OTUs to obtain integer values
(19, 20). Third, a variety of center log-ratio methods have been
developed where the compositional nature of the OTU counts is removed
and used to calculated Euclidean distances (aka Aitchison distances)
(18, 20--23). This strategy is purported to control for uneven
sequencing effort (22, 24); however, some have noticed that this feature
breaks down under certain conditions (25). Finally, as mentioned above,
variance stabilization transformations have been recommended to generate
values that can be used to calculate Euclidean distances (11).

The ongoing controversy over the use of rarefaction and the recent
development of alternative strategies to control for uneven sequencing
effort caused me to question how these methods compared to each other
using a simulation framework that overcame the issues with the McMurdie
and Holmes study. My analysis included 16S rRNA gene sequence data from
from 12 studies that characterized the variation in bacterial
communities from diverse environments (Table 1 and Figure S1). The
sequences were assigned to OTUs using a standard pipeline and their
frequencies and the number of sequences found in each sample were used
to generate simulated communities and treatment effects. For each
dataset and simulation, 100 replicate datasets were generated and used
as input to each of the strategies for controlling for uneven sequencing
effort. My overall conclusion was that rarefaction outperformed the
alternative strategies.

@@

\textbf{\emph{Without rarefaction, metrics of alpha diversity are
sensitive to sequencing effort.}} To test the sensitivity of various
approaches of measuring alpha diversity to sequencing effort, I
generated null models using OTU counts and sequencing depths from 12
studies. Under a null model, each community from the same dataset would
be expected to have the same alpha diversity regardless of the
sequencing effort. I measured the richness of the communities in each
dataset without any correction, using scaled ranked subsampling (SRS)
normalized OTU counts, with estimates based on non-parametric and
parametric approaches, and using rarefaction (e.g.~Figure S2). For each
dataset, all of the approaches, except for rarefaction, showed a strong
correlation between richness and the number of sequences in the sample
(Figure 1A). Next, I assessed diversity using the Shannon diversity
index and the inverse Simpson diversity index without any correction,
using normalized OTU counts, and rarefaction; I also used a
non-parametric estimator of Shannon diversity. The correlation between
sequencing depth and the diversity metric was not as strong as it was
for richness and the inverse Simpson diversity values were less
sensitive than the Shannon diversity values; however, the correlation to
the diversity metrics calculated with rarefaction were the lowest for
all of the metrics and studies (Figure 1A). The alpha-diversity metrics
calculated with rarefaction consistently demonstrated a lack of
sensitivity to sequencing depth.

\textbf{\emph{Without rarefaction, metrics of beta diversity are
sensitive to sequencing effort.}} To test the sensitivity of various
approaches of measuring beta diversity to sequencing effort, I used the
same null models used for studying the sensitivity of alpha diversity.
Under a null model, the ecological distance between any pair of samples
would be the same regardless of the difference in the number of
sequences observed in each sample (e.g., Figure S3). First, I calculated
the Jaccard distance coefficient between all pairs of communities within
a dataset. The Jaccard distance coefficient is the fraction of OTUs that
are unique to either community and does not account for the abundance of
the OTUs. Jaccard distances were calculated using the uncorrected OTU
counts, with rarefaction, relative abundances, and following
normalization using cumulative sum scaling (CSS) and SRS. Only the
distances calculated with rarefaction showed a lack of sensitivity to
sequencing effort (Figure 1B). Second, I analyzed the sensitivity of the
Bray-Curtis distance coefficient, which is a popular metric that
incorporates the abundance of each OTU. Similar to what I observed with
the Jaccard coefficient, only the data normalized with rarefaction
showed a lack of sensitivity to sequencing effort (Figure 1B). Third, I
calculated the Euclidean distance on raw OTU counts where the central
log-ratio (CLR) was calculated (i.e., Aitchison distances) by ignoring
OTUs in samples with zero counts (Robust CLR), adding a pseudo-count of
one to all OTU counts prior to calculating the CLR (One CLR), adding a
pseudo-count of one divided by the total number of sequences obtained
for the community (Nudge CLR), and imputing the value of zero counts
(Zero CLR). The Aitchison distances were all strongly sensitive to
sequencing effort (Figure 1B). Finally, I used the
variance-stabilization transformation (VST) from DESeq2 prior to
calculating Euclidean distances. Again, there was a strong sensitivity
to sequencing effort (Figure 1B). Although Euclidean distances are not
typically used on raw or rarefaction-normalized count data in ecology
(26), Euclidean distances calculated with rarefaction were not sensitive
to sequencing effort. Across each of the beta diversity metrics and
approaches used to account for uneven sequencing effort and sparsity,
rarefaction was the least sensitive approach to differences in
sequencing effort.

\textbf{\emph{Rarefaction limits the detection of false positives when
sequencing effort and treatment group are confounded.}} Next, I
investigated the impact of the various strategies and metrics on falsely
detecting a significant difference using the the same communities
generated from the null model in the analysis of alpha and beta
diversity metrics. To test for differences in alpha and beta diversity I
used the non-parametric Wilcoxon test and non-parametric
permutation-based multivariate analysis of variance (PERMANOVA). First,
I employed an unbiased null treatment model to measure the false
detection rate, which should not have meaningfully differed from 5\%.
Indeed, for each dataset and alpha and beta diversity metric and
strategy for accounting for uneven sequencing, approximately 5\% of the
tests yielded a significant result (Figure 2). Second, I employed a
biased null treatment model where the treatment group was completely
confounded with the number of sequences in each sample. Under this
model, only the data normalized with rarefaction consistently resulted
in a 5\% false positive rate for alpha and beta diversity metrics
(Figure 2). These results aligned with the observed sensitivity of alpha
and beta diversity metrics to sequencing effort and underscore the value
of rarefaction.

\textbf{\emph{Rarefaction preserves the statistical power to detect
differences between treatment groups.}} To assess the impact of
different approaches to control for uneven sequencing effort I performed
two additional simulations. In the first simulation, I implemented a
skewed abundance distribution model to create two treatment groups for
each dataset that were each populated with half of the samples each with
the same number of sequences as the samples had in the observed data.
The two treatment groups varied in their structure such that one had the
same abundances as the null distribution above and the other had 10\% of
its OTUs randomly selected to increase their counts by 5\%. The power to
detect differences in richness between the two simulated treatment
groups by all approaches was low (Figure 4A). This was likely because
the approach for generating the perturbed community did not necessarily
change the number of OTUs in each treatment group. Regardless, the
simulations testing differences in richness using the Rice and Stream
datasets had the greatest power when the richness data were calculated
with rarefaction. To explore this further, a richness-adjusted community
model was created by removing 3\% of the OTUs from a null model model.
As suggested by the first simulation, the richness data calculated with
rarefaction had a higher statistical power than the other approaches
when measuring richness (Figure 5). The simulations testing the power to
detect differences in Shannon diversity also showed that rarefaction
performed better than the other methods (Figure 4A). When testing for
differences in the Inverse Simpson diversity index the the difference
between rarefaction and the other methods was negligible (Figure 4A).
For tests of beta diversity I found that rarefaction was the most
reliable approach to maintain statistical power to detect differences
between two communities. Among the tests using the Jaccard and
Bray-Curtis metrics, raw count data and CSS normalized data had little
power relative to using rarefaction, relative abundance, and SRS to
normalize the uneven sequencing depths. The differences in power between
counts normalized with rarefaction, relative abundance, and SRS data was
small, but if there were differences, the power obtained using
rarefaction was greater than the other methods. Among the tests using
Euclidean distances, using raw counts and CLR and DESeq2 transformed
data had little power relative to the distances calculated using
rarefaction and relative abundances. This power-based analysis of the
simulated communities using different methods of handling uneven sample
sizes demonstrated the value of rarefaction for preserving the
statistical power to detect differences between treatment groups for
measures of alpha and beta diversity.

\textbf{\emph{Increased rarefaction depth reduces intra-sample variation
in alpha and beta diversity.}} Once concern with using rarefaction is
the perceived loss of sequencing information when a large fraction of
data appears to be removed when the community with the greatest
sequencing depth is sampled to the size of the community with the least
(e.g., the smallest sample in the Bioethanol dataset had 1.04\% of
sequences that were in the largest sample). To assess the sensitivity of
alpha and beta diversity metrics to rarefaction depth, I again used the
dataset generated using the null models, but used rarefaction with each
community to varying sequencing depths (Figure 6). The richness values
increased with sequencing effort as rare OTUs would continue be
detected. In contrast, the Shannon diversity and Bray-Curtis values
plateaued with increased sequencing effort. This result was expected
since increased sequencing would lead to increased precision in the
measured relative abundance of the OTUs. Next, I measured the
coefficient of variation (i.e., the mean divided by the standard
deviation) between samples for richness, Shannon diversity, and
Bray-Curtis distances. Although the mean richness appeared to increased
unbounded with sequencing effort, the coefficient of variation for each
dataset was relatively stable. In general, the coefficient of variation
increased slightly with sequencing depth only to decline once smaller
samples were removed from the analysis at higher sequencing depths.
Interestingly, the coefficient of variation between Shannon diversity
values decreased towards zero with increased sequencing effort and the
coefficient of variation between Bray-Curtis distances tended to
increase. Regardless, the coefficients of variation were relatively
small. This analysis indicates that there are benefits to increased
sequencing depths.

\textbf{\emph{Most studies have a high level of sequencing coverage.}}
To explore the concern over loss of sequencing depth further, I
calculated the Good's coverage for the observed data. The median
coverage for each dataset ranged between 89.4 and 99.8\% for the
Seagrass and Human datasets, respectively (Figure 7). When I used a
rarefaction threshold with each dataset at the size of the smallest
community in the dataset, with the exception of the Seagrass, Rice, and
Stream datasets, the median coverage with rarefaction was still greater
than 90\%. These results suggest that most studies had a level of
sequencing coverage that aligned with the diversity of the communities.
Next, I used the null model for each dataset to ask how much sequencing
effort was required to obtain higher levels of coverage. To obtain 95
and 99\% coverage, an average of 2.70 and 101.20-fold more sequence data
was estimated to be required than was required to obtain 90\% coverage,
respectively (Figure 7). As suggested by the simulated coverages curve
in Figure 7, these estimates are conservative. Regardless, the
sequencing effort required to achieve higher sequencing depth would
likely limit the number of samples that could be sequenced when
controlling for costs. Although it may be disconcerting to use
rarefaction to normalize to a sequencing depth that is considerably
lower than that obtained for the best sequenced community in a dataset,
sequencing coverage for many environments is probably adequate even at
the lower sequencing depth. Of course, regardless of the concerns
surrounding the choice of the rarefaction depth, the results throughout
this study demonstrate that rarefaction is necessary to avoid reaching
incorrect inferences.

@@

Over the past decade, the question of whether to use rarefaction with
microbial community sequence data has become controversial. The analyses
I presented here strongly indicate that rarefaction is necessary to
control for uneven sequencing effort when comparing communities based on
alpha and beta diversity indices. Compared to all other methods,
rarefaction removed the correlation between sequencing depth and alpha
or beta diversity metrics when the sequencing depth varied by as much as
97-fold across samples. I showed that this correlation could lead
falsely detecting differences between treatment groups if sequencing
depth and sequencing effort are confounded. The correlation with
sequencing effort leads to an artificial increase in the variation
between samples and a reduced power to detect true differences in alpha
and beta diversity. For these reasons, rarefaction is a valuable tool to
control for uneven sequencing effort until improved statistical
procedures are developed or it becomes possible to more evenly
distribute sequencing effort across samples.

The primary alternative to rarefaction for measuring alpha diversity is
to estimate the metric using non-parametric or parametric models with
raw counts and to then compare the estimates (7, 14). My results
demonstrate that such approaches are limited for several reasons. First,
non-parametric richness estimates such as ACE and Chao1 are sensitive to
sequencing effort. Therefore, these strategies do not, in practice,
remove the effects of sequencing effort. Second, parametric approaches,
such as those implemented in the \texttt{breakaway} R package, generate
confidence intervals that are likely to include the true richness and
theoretically shrink with increased sequencing effort. Yet for most
samples, the confidence intervals around the estimates are too wide to
be meaningful, again leading to an inability to remove the effects of
sequencing effort. Third, it has become an increasingly common practice
for researchers to remove sequences that are rare in a sample (e.g.,
those that appear once). Although that approach was not taken in this
study, removing rare sequences would skew the distribution of sequences
and OTUs leading to a distortion of the measurement of alpha diversity
(27). The effects of removing rare sequences would vary across samples
depending on the number of sequences in each sample. One interesting
result of this analysis was the demonstration that as metrics that
depend less on rare taxa are used, the effect of uneven sequencing
effort was reduced. For example, richness counts a sequence appearing
once as much as sequence appearing 1,000 times, while the Shannon
diversity index places more emphasis on the more abundant sequence, and
the inverse Simpson index even more. Although normalizing communities to
a common number of sequences is also suggested (e.g., SRS normalization)
to control for uneven sequencing effort, the current analysis indicates
that its performance does not meet that of rarefaction. For analysis of
alpha diversity metrics, rarefaction is the most effective and
consistent approach to control for uneven sequencing effort.

Use of relative abundances, normalized counts, variance stabilizing
transformations, and centered log ratios have each been recommended as
more robust alternatives to rarefaction. Again, the only approach that
consistently removed the effects of uneven sequencing effort was
rarefaction. Most of the recommendations borrow techniques from methods
for identifying differential gene expression. Unfortunately, there is an
important but under appreciated difference between gene expression and
community data. This is the interpretation of unobserved data. For gene
expression analysis in a single organism the lack of any sequencing data
for a gene would indicate that although the gene was present, its
expression was below the limit of detection. Sequencing the same
organism under multiple conditions would not lead to a seemingly
unbounded number of genes in the organisms. Rather, the number of genes
has a definite limit that is knowable from the genome sequence. With
microbiome data, an unobserved sequence could mean that the organism was
present, but below the limit of detection or that the organism was
missing. Because we have yet to exhaustively sample any community in the
same way we have sequenced a single genome, it is unreliable to impute
the presence of all organisms. Yet, this is exactly what the variance
stabilization transformation and most CLR techniques do. This analysis
has demonstrated a clear correlation between distances calculated by
these methods and sequencing effort. This result is at odds with the
claims by others that the distances are scale invariant (22, 24). Again,
rarefaction is the most effective and consistent approach to control for
uneven sequencing effort when calculating beta diversity metrics.
